---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# bipl5

<!-- badges: start -->
[![R-CMD-check](https://github.com/RuanBuys/bipl5/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/RuanBuys/bipl5/actions/workflows/R-CMD-check.yaml)
[![Codecov test coverage](https://codecov.io/gh/RuanBuys/bipl5/branch/main/graph/badge.svg)](https://app.codecov.io/gh/RuanBuys/bipl5?branch=main)
[![pkgcheck](https://github.com/RuanBuys/bipl5/workflows/pkgcheck/badge.svg)](https://github.com/RuanBuys/bipl5/actions?query=workflow%3Apkgcheck)
<!-- badges: end -->

The goal of bipl5 is to provide a modern take on PCA biplots with calibrated axes. The biplots are rendered in HTML via the plotly graphing library, upon which custom JavaScript code is appended to make the plot reactive. The traditional biplot view is also extended through an algorithm that translates the axes out of the data centroid, decluttering the view. In addition to the former, inter-class kernel densities are superimposed on the axes to give an indication of the data spread across the variables.

## Installation

You can install the latest version of bipl5 from CRAN with:

```{r}
install.packages("bipl5")
library(bipl5)
```

## Example: PCA biplot

The current workflow is built around `biplotEZ` objects wrapped with `wrap_bipl5()`. The resulting widget includes the calibrated axes, translated density axes, and the interactive controls:

```{r example}
bp <- biplotEZ::biplot(iris[, -5]) |>
  biplotEZ::PCA() |>
  wrap_bipl5()

plot(bp)
```

## Example: CVA biplot

Canonical variate biplots use the same workflow:

```{r}
bp_cva <- biplotEZ::biplot(iris[, 1:4]) |>
  biplotEZ::CVA(classes = iris[, 5]) |>
  wrap_bipl5()

plot(bp_cva)
```

Translated axes are part of the interactive widget. After rendering the plot, use the translated-axis controls in the plot UI rather than calling a separate translation function.
