---
title: "PCA Biplots"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{PCA Biplots}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 8,
  fig.height = 6,
  out.width = "100%"
)
options(crayon.enabled = FALSE)
```

```{r setup}
#suppressPackageStartupMessages(library(bipl5))
library(devtools)
load_all()
```

This vignette focuses on the modern `bipl5` workflow for principal component
analysis (PCA) biplots:

```r
init_biplot(...) |> scale_mds(type = "pca", ...)
```

The mathematical background is the same background documented in
`wrap_bipl5.PCA()`, but the examples below use the newer specification-based
workflow because it is the most natural entry point for package users who want
to build a single display first and then extend it later.

The goals of this vignette are to:

- construct a PCA biplot with `init_biplot()` and `scale_mds()`
- show the printed structure of the resulting `bipl5_biplot`
- add extra `mdsDisplays` with `append_mdsDisplay()`
- demonstrate `format_samples()` with both one and two categorical
  stratifications
- explain the PCA fit measures stored in `fit_measures`
- preserve the current long-form documentation of the PCA biplot mathematics

# Building a PCA biplot

`init_biplot()` stores the data and preprocessing choices, while `scale_mds()`
fits and compiles the requested biplot. If the input data frame contains
categorical columns, those columns are retained inside the specification object
so they can be used later by `format_samples()`.

```{r data-and-build}
iris2 <- iris
iris2$Band <- factor(
  rep(c("class1", "class2", "class3", "class4"), length.out = nrow(iris2))
)

pca_spec <- init_biplot(iris2, scale = TRUE)

bp_pca <- pca_spec |>
  scale_mds(
    type = "pca",
    eigenvectors = c(1, 2),
    group_aes = iris2$Species,
    show_group_means = TRUE
  )
```

Immediately after `scale_mds()`, the object contains one compiled display and a
PCA fit-measures branch:

```{r print-single}
bp_pca
```

The default PCA plot is interactive. The left-hand side is the biplot itself,
while the right-hand side fit panel can be opened from the widget controls.

```{r plot-single}
plot(bp_pca)
```

# Adding more mdsDisplays

The `scale_mds()` workflow deliberately compiles only the requested pair of
principal components. This keeps the initial object light, and it mirrors the
fact that a user often wants to start with just one view. Additional views can
then be added with `append_mdsDisplay()`.

Here we extend the original `PC 1 & 2` display to also include `PC 1 & 3` and
`PC 2 & 3`:

```{r append-displays}
bp_pca_full <- bp_pca |>
  append_mdsDisplay(c(1, 3)) |>
  append_mdsDisplay(c(2, 3))
```

The printed tree now reflects three stored displays and one fit table for each
pair:

```{r print-multi}
bp_pca_full
```

The plot now exposes a dropdown for switching between the stored PC pairs.

```{r plot-multi}
plot(bp_pca_full)
```

# Formatting samples

`format_samples()` does not refit the PCA model. It rebuilds the sample-trace
block inside each stored `mdsDisplay`, so it is the right tool when the
ordination is fixed but the display grouping needs to change.

We begin with a colour stratification by species:

```{r format-color}
bp_pca_species <- bp_pca_full |>
  format_samples(
    stratify = "col",
    by = Species,
    col = c("tomato", "steelblue", "darkgreen")
  )
```

We then add a second categorical stratification through plotting symbols. Since
`Band` is different from `Species`, the plotted observation traces are rebuilt
internally as the observed `Species x Band` combinations, while the legend is
kept readable by showing separate legend sections for the colour and symbol
variables.

```{r format-dual}
bp_pca_dual <- bp_pca_species |>
  format_samples(
    stratify = "symbol",
    by = Band,
    pch = c(15, 16, 17, 18)
  )
```

The formatted object remains a `bipl5_biplot`, so it can still be printed,
subset, extended, and plotted as usual.

```{r plot-dual}
plot(bp_pca_dual)
```

# PCA fit measures

PCA biplots are the richest objects in the current package because they carry a
full `fit_measures` branch. The print method makes the available fit objects
explicit:

```{r print-fit-measures}
bp_pca_full$fit_measures
```

In the interactive PCA widget, these are exposed through the "Measures of Fit"
button and the fit-panel dropdown. The fit objects are:

- `CumPred`: cumulative predictivity across increasing subspace dimension. Each
  line tracks one variable's cumulative axis predictivity, and the weighted
  mean line equals the overall display quality.
- `CumAd`: cumulative adequacy across increasing subspace dimension. This shows
  how well each variable direction is represented by the accumulated loading
  structure, irrespective of the final two-dimensional display pair.
- `VarExp`: variance explained. In the current implementation this combines a
  line for the individual proportion of variance explained by each principal
  component with stacked variable-level contributions.
- `Scree`: the eigenvalue profile across the principal components.
- `fit_table_ij`: the pair-specific summary table for the active display. Each
  table reports variable-level adequacy and predictivity for that specific
  `PC i & j` map.

If you want one of the graph-based fit objects on its own, you can extract it
and plot it directly:

```{r fit-graph-example}
fit_scree <- extract(bp_pca_full, fit_measures, Scree)
plot(fit_scree)
```

The summary table is not a separate `bipl5_fit` graph object, because it is
already stored in the `fit_measures` branch as a plotly table trace. It updates
when the active `mdsDisplay` changes.

# Mathematical background

This section preserves the current long-form PCA documentation in narrative
form.

Let $\mathbf{X} \in \mathbb{R}^{n \times p}$ denote the processed data matrix
stored in the input object after centring and any optional scaling. PCA is
performed on that processed matrix, so the singular value decomposition is

$$
\mathbf{X} = \mathbf{U}\mathbf{D}\mathbf{V}^{\top},
$$

where $\mathbf{U}^{\top}\mathbf{U} = \mathbf{I}$,
$\mathbf{V}^{\top}\mathbf{V} = \mathbf{I}$, and
$\mathbf{D} = \operatorname{diag}(d_1, \ldots, d_q)$ with
$q = \operatorname{rank}(\mathbf{X})$ and
$d_1 \ge \cdots \ge d_q \ge 0$. The principal component score vectors are

$$
\mathbf{z}_t = d_t \mathbf{u}_t = \mathbf{X}\mathbf{v}_t,
\qquad t = 1, \ldots, q.
$$

Suppose the displayed pair is $(a, b)$ with $a < b$. Let
$\mathbf{J}_{ab}$ be the diagonal selector matrix with ones in positions $a$
and $b$. Then the two-dimensional fitted matrix is

$$
\widehat{\mathbf{X}}_{ab} =
\mathbf{U}\mathbf{D}\mathbf{J}_{ab}\mathbf{V}^{\top}
=
\mathbf{X}\mathbf{V}\mathbf{J}_{ab}\mathbf{V}^{\top}.
$$

Equivalently, with $\mathbf{U}_{ab}$, $\mathbf{D}_{ab}$, and
$\mathbf{V}_{ab}$ containing only the selected components,

$$
\widehat{\mathbf{X}}_{ab}
=
\mathbf{U}_{ab}\mathbf{D}_{ab}\mathbf{V}_{ab}^{\top}
=
d_a \mathbf{u}_a \mathbf{v}_a^{\top}
+
d_b \mathbf{u}_b \mathbf{v}_b^{\top}.
$$

For the ordinary PCA biplot the displayed sample coordinates are

$$
\mathbf{Z}_{ab} = \mathbf{U}_{ab}\mathbf{D}_{ab},
\qquad
\mathbf{H}_{ab} = \mathbf{V}_{ab},
$$

so that

$$
\widehat{\mathbf{X}}_{ab} = \mathbf{Z}_{ab}\mathbf{H}_{ab}^{\top}.
$$

If $\mathbf{h}_{(j)}$ is the displayed direction for variable $j$, then the
fitted value for observation $i$ on variable $j$ is

$$
\widehat{x}_{ij} = \mathbf{z}_i^{\top}\mathbf{h}_{(j)},
$$

and the calibrated axis point corresponding to marker value $\mu$ is

$$
\mathbf{p}_{\mu j}
=
\frac{\mu}{\mathbf{h}_{(j)}^{\top}\mathbf{h}_{(j)}}\mathbf{h}_{(j)}.
$$

For a correlation biplot the factorization can instead be written as

$$
\widehat{\mathbf{X}}_{ab} =
\mathbf{U}_{ab}(\mathbf{V}_{ab}\mathbf{D}_{ab})^{\top},
$$

so the displayed sample coordinates are $\mathbf{U}_{ab}$ and the displayed
variable directions are the rows of $\mathbf{V}_{ab}\mathbf{D}_{ab}$. In that
setting the variable directions are tuned to variable-correlation structure
rather than raw score geometry.

Two orthogonality relations are central:

$$
\mathbf{X}\mathbf{X}^{\top}
=
\widehat{\mathbf{X}}_{ab}\widehat{\mathbf{X}}_{ab}^{\top}
+
\mathbf{E}_{ab}\mathbf{E}_{ab}^{\top},
$$

and

$$
\mathbf{X}^{\top}\mathbf{X}
=
\widehat{\mathbf{X}}_{ab}^{\top}\widehat{\mathbf{X}}_{ab}
+
\mathbf{E}_{ab}^{\top}\mathbf{E}_{ab},
$$

where $\mathbf{E}_{ab} = \mathbf{X} - \widehat{\mathbf{X}}_{ab}$. The first
justifies sample-side fit measures (Type A orthogonality), and the second
justifies variable-side fit measures (Type B orthogonality).

For sample $i$, the sample predictivity is

$$
\psi_i
=
\frac{\|\widehat{\mathbf{x}}_{i\cdot}\|^2}
     {\|\mathbf{x}_{i\cdot}\|^2}
=
1 -
\frac{\|\mathbf{x}_{i\cdot} - \widehat{\mathbf{x}}_{i\cdot}\|^2}
     {\|\mathbf{x}_{i\cdot}\|^2}.
$$

For variable $j$, the axis predictivity is

$$
\phi_j
=
\frac{\|\widehat{\mathbf{x}}_{(j)}\|^2}
     {\|\mathbf{x}_{(j)}\|^2}
=
1 -
\frac{\|\mathbf{x}_{(j)} - \widehat{\mathbf{x}}_{(j)}\|^2}
     {\|\mathbf{x}_{(j)}\|^2}.
$$

The overall quality of the displayed PCA plane is

$$
R^2_{\mathrm{disp},ab}
=
\frac{\|\widehat{\mathbf{X}}_{ab}\|_F^2}{\|\mathbf{X}\|_F^2}
=
\frac{d_a^2 + d_b^2}{\sum_{t = 1}^q d_t^2}.
$$

Because of orthogonality, this same quantity can be read as a weighted average
of either the sample predictivities or the axis predictivities. On the
variable side,

$$
R^2_{\mathrm{disp},ab}
=
\sum_{j = 1}^p w_j \phi_j,
\qquad
w_j = \frac{\|\mathbf{x}_{(j)}\|^2}{\|\mathbf{X}\|_F^2}.
$$

When the processed variables all have equal sums of squares, the overall
quality is simply the average axis predictivity. This is one reason
standardized PCA is often easy to interpret.

PCA also gives a clean per-component decomposition. Since

$$
\widehat{\mathbf{X}}_{ab}
=
d_a \mathbf{u}_a \mathbf{v}_a^{\top}
+
d_b \mathbf{u}_b \mathbf{v}_b^{\top},
$$

the two displayed rank-1 pieces are orthogonal, and therefore

$$
R^2_{\mathrm{disp},ab} = R_a^2 + R_b^2,
\qquad
R_a^2 = \frac{d_a^2}{\|\mathbf{X}\|_F^2},
\qquad
R_b^2 = \frac{d_b^2}{\|\mathbf{X}\|_F^2}.
$$

Likewise,

$$
\phi_j = \phi_{ja} + \phi_{jb},
\qquad
\psi_i = \psi_{ia} + \psi_{ib},
$$

so both axis-side and sample-side quality can be broken down into the two
displayed components.

The package documentation also distinguishes these sum-of-squares measures from
the direct-reading diagnostics of Alves (2012). If $\mathbf{g}_{(j)}$ is the
displayed direction of variable $j$, then the read-off value on that axis for
sample $i$ is

$$
\widehat{x}_{ij} = \mathbf{z}_i^{\top}\mathbf{g}_{(j)}.
$$

The direct-reading error is

$$
\delta_{ij} = \frac{|x_{ij} - \widehat{x}_{ij}|}{s_j},
$$

where $s_j$ is the standard deviation used in preprocessing, and the mean
axis-level direct-reading error is

$$
\bar{\delta}_j = \frac{1}{n}\sum_{i = 1}^n \delta_{ij}.
$$

These quantities answer a different question from predictivity. Predictivity
asks how much sum of squares is reproduced by the displayed plane. The Alves
diagnostics ask how accurately values can be read directly from the plotted
calibrated axes.

The per-observation direct-reading errors $\delta_{ij}$ can be attached to a
biplot with `score_axes()`. Inserting this step into the construction pipeline

```r
init_biplot(data, scale = TRUE) |>
  scale_mds(type = "pca") |>
  score_axes() |>
  plot()
```

adds an extra `Error` column, expressed as a percentage ($100\,\delta_{ij}$),
to the hover table that appears when a data point is hovered over, alongside its
actual and predicted values. The calculation is performed for every displayed
dimension pair (`mdsDisplay`) in the object, and applies equally to PCA, PCO and
CVA biplots. Because the diagnostic is opt-in, it is only computed when
`score_axes()` is called.

# References

Alves, M. R. (2012). Evaluation of the predictive power of biplot axes to
automate the construction and layout of biplots based on the accuracy of
direct readings from common outputs of multivariate analyses: application to
principal component analysis. *Journal of Chemometrics*, 26(5), 180-190.

Eckart, C. and Young, G. (1936). The approximation of one matrix by another of
lower rank. *Psychometrika*, 1, 211-218.

Gabriel, K. R. (1971). The biplot graphical display of matrices with
application to principal component analysis. *Biometrika*, 58(3), 453-467.

Gardner-Lubbe, S., le Roux, N. J. and Gower, J. C. (2008). Measures of fit in
principal component and canonical variate analyses. *Journal of Applied
Statistics*, 35(9), 947-965.

Gower, J. C. and Hand, D. J. (1996). *Biplots*. London: Chapman and Hall.

Gower, J. C., Lubbe, S. and le Roux, N. J. (2011). *Understanding Biplots*.
Chichester: Wiley.

Greenacre, M. (2010). *Biplots in Practice*. Bilbao: BBVA Foundation.
