---
title: "Regression Biplots"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Regression Biplots}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 8,
  fig.height = 6,
  out.width = "100%"
)
options(crayon.enabled = FALSE)
```

```{r setup}
suppressPackageStartupMessages(library(bipl5))
```

This vignette covers regression biplots constructed with the newer `bipl5`
workflow:

```r
init_biplot(...) |> scale_mds(type = "regress", Z = ...)
```

The mathematics mirror the current `wrap_bipl5.regress()` documentation, but
the examples use `init_biplot()` and `scale_mds()` because this is now the most
direct way to build a regression biplot inside the package.

Regression biplots differ from PCA biplots in one crucial way: the sample map
is supplied from outside. The display coordinates in `Z` are treated as fixed,
and the variable axes are then fitted to that map by least squares.

# Building a regression biplot

As in the PCA vignette, we keep an extra categorical column in the original
data so we can later demonstrate `format_samples()` with a second sample
stratification.

```{r build-regression}
iris2 <- iris
iris2$Band <- factor(
  rep(c("class1", "class2", "class3", "class4"), length.out = nrow(iris2))
)

Z <- prcomp(iris2[, 1:4], scale. = TRUE)$x[, 1:2]

reg_spec <- init_biplot(iris2, scale = TRUE)

bp_reg <- reg_spec |>
  scale_mds(
    type = "regress",
    Z = Z,
    group_aes = iris2$Species,
    show_group_means = TRUE
  )
```

After `scale_mds(type = "regress", ...)`, the object contains one fixed display
only:

```{r print-regression}
bp_reg
```

That single display is always stored as `mdsDisplay_12` and labelled
`Dim 1 & 2`, because the geometry comes from the supplied two-column map
instead of an internally generated sequence of principal components.

```{r plot-regression}
plot(bp_reg)
```

# Formatting samples

`format_samples()` works for regression biplots in the same way that it works
for PCA biplots: the fitted regression geometry is left untouched, and only the
sample-trace block is rebuilt.

First we colour the samples by species:

```{r format-regression-color}
bp_reg_species <- bp_reg |>
  format_samples(
    stratify = "col",
    by = Species,
    col = c("tomato", "steelblue", "darkgreen")
  )
```

Then we add a second categorical stratification by plotting symbol:

```{r format-regression-dual}
bp_reg_dual <- bp_reg_species |>
  format_samples(
    stratify = "symbol",
    by = Band,
    pch = c(15, 16, 17, 18)
  )
```

The resulting widget now has separate sample legend sections for `Species` and
`Band`, while the hidden observation traces are the observed `Species x Band`
combinations.

```{r plot-regression-dual}
plot(bp_reg_dual)
```

# How regression biplots differ from PCA biplots

Regression biplots are intentionally simpler in the current package:

- they have one fixed `mdsDisplay` only
- `append_mdsDisplay()` and `remove_mdsDisplay()` are not supported
- they do not carry the PCA `fit_measures` branch
- there is no PCA-style right-hand fit panel
- the fitted variable axes are tied to the supplied map `Z`

The main quality label in the widget still reports the overall
regression-biplot quality and its ordered dimension-specific contributions, but
the multi-display PCA fit machinery is absent by design.

# Mathematical background

This section preserves the long-form regression-biplot documentation in a
vignette-style narrative.

Let $\mathbf{X} \in \mathbb{R}^{n \times p}$ denote the processed data matrix
after centring and optional scaling, and let
$\mathbf{Z} \in \mathbb{R}^{n \times 2}$ denote the externally supplied display
coordinates. In a regression biplot the sample map is taken as given, and the
variable axes are fitted to that map by least squares.

The fitted model is

$$
\mathbf{X} = \mathbf{Z}\mathbf{H}^{\top} + \mathbf{E},
$$

with

$$
\mathbf{H}^{\top} = (\mathbf{Z}^{\top}\mathbf{Z})^{-1}\mathbf{Z}^{\top}\mathbf{X}
$$

when $\mathbf{Z}$ has full column rank. Equivalently,

$$
\widehat{\mathbf{X}}
=
\mathbf{Z}\mathbf{H}^{\top}
=
\mathbf{Z}(\mathbf{Z}^{\top}\mathbf{Z})^{-1}\mathbf{Z}^{\top}\mathbf{X}
=
\mathbf{P}_Z \mathbf{X},
$$

where $\mathbf{P}_Z$ is the orthogonal projector onto the column space of
$\mathbf{Z}$.

If $\mathbf{h}_{(j)}$ is the fitted direction for variable $j$, then the
predicted value for sample $i$ on that variable is

$$
\widehat{x}_{ij} = \mathbf{z}_i^{\top}\mathbf{h}_{(j)}.
$$

The point on the calibrated axis corresponding to marker value $\mu$ is

$$
\mathbf{p}_{\mu j}
=
\frac{\mu}{\mathbf{h}_{(j)}^{\top}\mathbf{h}_{(j)}}\mathbf{h}_{(j)}.
$$

So, just as in the PCA case, direct readings from the axis coincide with the
fitted values from the display.

Because $\widehat{\mathbf{X}} = \mathbf{P}_Z \mathbf{X}$ is an orthogonal
projection, the regression biplot admits a natural variable-side predictivity
measure. For variable $j$,

$$
\phi_j
=
\frac{\|\widehat{\mathbf{x}}_{(j)}\|^2}
     {\|\mathbf{x}_{(j)}\|^2}
=
1 -
\frac{\|\mathbf{x}_{(j)} - \widehat{\mathbf{x}}_{(j)}\|^2}
     {\|\mathbf{x}_{(j)}\|^2}.
$$

This is the proportion of variable $j$'s sum of squares reproduced by the
display. It is also the ordinary multiple-regression $R^2$ obtained by
regressing that variable on the supplied coordinates in $\mathbf{Z}$.

The overall regression-biplot quality is

$$
R^2_{\mathrm{disp}}
=
\frac{\|\widehat{\mathbf{X}}\|_F^2}{\|\mathbf{X}\|_F^2}
=
1 -
\frac{\|\mathbf{X} - \widehat{\mathbf{X}}\|_F^2}{\|\mathbf{X}\|_F^2}.
$$

Because the column-wise decomposition is orthogonal, the same quantity can be
written as a weighted average of the variable predictivities:

$$
R^2_{\mathrm{disp}}
=
\sum_{j = 1}^p w_j \phi_j,
\qquad
w_j = \frac{\|\mathbf{x}_{(j)}\|^2}{\|\mathbf{X}\|_F^2}.
$$

If the processed variables all have equal sums of squares, the overall quality
is simply the average of the variable predictivities.

To separate the quality into contributions from the two displayed dimensions,
the package uses an ordered orthogonalization of the supplied coordinates. Let

$$
\mathbf{u}_1 = \mathbf{z}_1,
\qquad
\mathbf{q}_1 = \frac{\mathbf{u}_1}{\|\mathbf{u}_1\|},
$$

and then let

$$
\mathbf{u}_2 =
\mathbf{z}_2 - \mathbf{q}_1 \mathbf{q}_1^{\top}\mathbf{z}_2,
\qquad
\mathbf{q}_2 = \frac{\mathbf{u}_2}{\|\mathbf{u}_2\|}.
$$

Equivalently, this is the QR orthogonalization of $\mathbf{Z}$ while
preserving the supplied column order. The display projector can then be written
as

$$
\mathbf{P}_Z = \mathbf{Q}\mathbf{Q}^{\top},
\qquad
\mathbf{Q} = [\mathbf{q}_1 \ \mathbf{q}_2].
$$

Hence

$$
\widehat{\mathbf{X}}
=
\mathbf{q}_1 \mathbf{q}_1^{\top}\mathbf{X}
+
\mathbf{q}_2 \mathbf{q}_2^{\top}\mathbf{X},
$$

and the two pieces are orthogonal. This yields the ordered contributions

$$
R_1^2 =
\frac{\|\mathbf{q}_1 \mathbf{q}_1^{\top}\mathbf{X}\|_F^2}{\|\mathbf{X}\|_F^2},
\qquad
R_{2 \mid 1}^2 =
\frac{\|\mathbf{q}_2 \mathbf{q}_2^{\top}\mathbf{X}\|_F^2}{\|\mathbf{X}\|_F^2},
$$

with

$$
R^2_{\mathrm{disp}} = R_1^2 + R_{2 \mid 1}^2.
$$

This decomposition is ordered. If the original columns of $\mathbf{Z}$ are not
orthogonal, then $R_{2 \mid 1}^2$ is the additional contribution of the second
supplied dimension after removing overlap with the first. It should not be
interpreted as the contribution of the raw second column in isolation.

The same idea gives an ordered decomposition of each variable predictivity:

$$
\phi_j = \phi_{j1} + \phi_{j, 2 \mid 1}.
$$

This is a useful way to understand which part of the supplied map is doing the
predictive work for each variable.

The documentation also preserves the distinction between predictivity and
direct-reading diagnostics. If $s_j$ is the standard deviation used in
preprocessing, the pointwise direct-reading error is

$$
\delta_{ij} = \frac{|x_{ij} - \widehat{x}_{ij}|}{s_j},
$$

and the mean direct-reading error for axis $j$ is

$$
\bar{\delta}_j = \frac{1}{n}\sum_{i = 1}^n \delta_{ij}.
$$

These are display-specific diagnostics. They measure how accurately a user can
read values directly from the currently drawn calibrated axis. They are not the
same thing as the variance-accounted-for quantities $\phi_j$ and
$R^2_{\mathrm{disp}}$.

One final distinction from PCA is especially important: a regression biplot
does not, in general, satisfy the sample-side orthogonality that would justify
PCA-style sample predictivities. So the principled sum-of-squares fit measures
here are the variable predictivities, the overall display quality, and the
ordered dimension-specific contributions derived from the supplied map.

# References

Alves, M. R. (2012). Evaluation of the predictive power of biplot axes to
automate the construction and layout of biplots based on the accuracy of
direct readings from common outputs of multivariate analyses: application to
principal component analysis. *Journal of Chemometrics*, 26(5), 180-190.

Gabriel, K. R. (1971). The biplot graphical display of matrices with
application to principal component analysis. *Biometrika*, 58(3), 453-467.

Gower, J. C. and Hand, D. J. (1996). *Biplots*. London: Chapman and Hall.

Gower, J. C., Lubbe, S. and le Roux, N. J. (2011). *Understanding Biplots*.
Chichester: Wiley.

Greenacre, M. (2010). *Biplots in Practice*. Bilbao: BBVA Foundation.

la Grange, A., le Roux, N. and Gardner-Lubbe, S. (2009). BiplotGUI:
Interactive Biplots in R. *Journal of Statistical Software*, 30(12), 1-37.
